---
title: "Atwaters siRNA Normalization Analysis"
output: html_document
date: '`r Sys.Date()`'
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,
                      echo=FALSE, warning=FALSE, message=FALSE)
```

```{r setup}
library("ggplot2")
library("data.table")
library("MEMA")
library("grid")
library("knitr")
library("gplots")
library("RColorBrewer")
library("DT")
library("d3heatmap")
library("plotly")
library(scales)

source("WellPlateFunctions.R")

#Calculate SSMD
SSMDPlate <- function(df){
  if(length(unique(df)))
    mup <- mean(df$WellCellCount[df$GeneSymbol=="PLK1"], na.rm = TRUE)
  sdp <- sd(df$WellCellCount[df$GeneSymbol=="PLK1"], na.rm = TRUE)
  mun <- mean(df$WellCellCount[df$GeneSymbol=="NegCtrl"], na.rm = TRUE)
  sdn <- sd(df$WellCellCount[df$GeneSymbol=="NegCtrl"], na.rm = TRUE)
  SSMD <- (mup-mun)/sqrt(sdp^2+sdn^2)
}

#Calculate SSMD on the loess normed well cell count
SSMDPlateLoess <- function(df){
  if(length(unique(df)))
    mup <- mean(df$WellCellCountLoessNorm[df$GeneSymbol=="PLK1"], na.rm = TRUE)
  sdp <- sd(df$WellCellCountLoessNorm[df$GeneSymbol=="PLK1"], na.rm = TRUE)
  mun <- mean(df$WellCellCountLoessNorm[df$GeneSymbol=="NegCtrl"], na.rm = TRUE)
  sdn <- sd(df$WellCellCountLoessNorm[df$GeneSymbol=="NegCtrl"], na.rm = TRUE)
  SSMD <- (mup-mun)/sqrt(sdp^2+sdn^2)
}

calcZPrimeLoess <- function(dt, method="original", posCtrl="PLK1"){
  imageHTS::zprime(dt[GeneSymbol=="NegCtrl",WellCellCountLoessNorm],dt[GeneSymbol==posCtrl,WellCellCountLoessNorm], method)
}

wDT <- fread("AtwatersWell_v1.1.txt")
wDT$OmeroDetailURL <- paste0('<a href="https://omero.ohsu.edu/webclient/img_detail/',wDT$ImageID,'/"',' target="_blank">Omero</a>')
wDT$OmeroThumbnailURL <- paste0('<a href="https://omero.ohsu.edu/webclient/render_thumbnail/',wDT$ImageID,'/"',' target="_blank">Omero</a>')
wDT$OmeroImageURL <- paste0('<a href="https://omero.ohsu.edu/webclient/render_image/',wDT$ImageID,'/"',' target="_blank">Omero</a>')

```

##Filtering
Data from the wells listed below have been removed from all analysis in this report.  

AT384P10037 (ZR75B): rows C and D have very low cell counts  
AT384P20038 (ZR75B): rows C and D have very low cell counts  
AT384P10049 (HCC38): columns 12 and 13 have missing cells  


```{r filter, echo=FALSE}
#Manually filter out selected rows of poor quality
wDT <- wDT[!(grepl("AT384P10037",wDT$Barcode) & grepl("^3$|^4$",wDT$Row)),]
wDT <- wDT[!(grepl("AT384P20038",wDT$Barcode) & grepl("^3$|^4$",wDT$Row)),]
wDT <- wDT[!(grepl("AT384P10049",wDT$Barcode) & grepl("^12$|^13$",wDT$Column)),]

```


```{r loessNorm, eval=TRUE}

span=.05
colNames <- grep("Omero|Image|Field",colnames(wDT),value=TRUE,invert=TRUE)

wDT <- unique(wDT[,colNames, with=FALSE])
wDT <- loessNormalize(wDT, signalNames="WellCellCount", span=span)

```

<br>

###Loess Normalization

Assume the model for each gene response is: 

$$y = y_{signal} + y_{unwanted~variation} + ε$$

The residual for the gene in row i and column j is calculated by: 

$$y_{r(i,j)} = y_{g(i,j)} - median_{g(i,j)}$$

where $y_{g(i,j)}$ is the raw signal value for the gene at row i and column j and $median_{g(i,j)}$ is the median of its replicates within the plate. 

Using the fitted bi-variate loess value of the residual at row i and column j $y_{rl(i,j)}$ with a span of `r span` to estimate the unwanted variation and subtracting this from the raw signal value yields the normalized signal.

$$y_{s(i,j)} = y_{g(i,j)} - y_{rl(i,j)}$$

Plate maps and boxplots of each step in the normalization for every plate in the dataset are shown in the supplemental material at the end of the report.  

##QA Metrics
The Z Prime Factor and SSMD QA metrics are calculated on the raw and the loess normalized Well Cell Count values. A span of `r span` was chosen for the loess normalization after performing the span analysis shown in the supplemental material.  

```{r calcQAMetrics}

#Remove rows from multiple fields/images of each well
colNames <- grep("Omero|Image|Field",colnames(wDT),value=TRUE,invert=TRUE)
DT <- unique(wDT[,colNames, with=FALSE], by=c("Barcode","Well"))

DT <- DT[,ZPrimeRobust := calcZPrime(.SD, method="robust"), by="Barcode", .SDcols=c("GeneSymbol","WellCellCount")]
DT <- DT[,SSMD := SSMDPlate(.SD), by="Barcode", .SDcols=c("GeneSymbol","WellCellCount")]
DT <- DT[,ZPrimeRobustLoess := calcZPrimeLoess(.SD, method="robust"), by="Barcode", .SDcols=c("GeneSymbol","WellCellCountLoessNorm")]
DT <- DT[,SSMDLoess := SSMDPlateLoess(.SD), by="Barcode", .SDcols=c("GeneSymbol","WellCellCountLoessNorm")]

DT <- unique(DT[,list(Barcode,ZPrimeRobust,SSMD,ZPrimeRobustLoess,SSMDLoess)])
wDT <- merge(wDT,DT, by="Barcode")

```

<br>

####Plates Passing QA on Raw Data  
Using the raw signal values, there are `r sum(DT$ZPrimeRobust>=0)` plates with Z Prime Robust factor >= 0, `r sum(DT$SSMD <= -3, na.rm=TRUE)` plates with SSMD <= -3 and `r sum(DT$SSMD <= -2, na.rm=TRUE)` plates with SSMD <= -2.

<br>

####Plates Passing QA on Normalized Data  
Using the a span of `r span` and the loess normalization, there are `r sum(DT$ZPrimeRobustLoess>=0)` plates with Z Prime Robust factor >= 0, `r sum(DT$SSMDLoess <= -3, na.rm=TRUE)` plates with SSMD <= -3 and `r sum(DT$SSMDLoess <= -2, na.rm=TRUE)` plates with SSMD <= -2.

The cell lines and plates passing QA with Z Prime factor values greater than or equal to 0 are:
`r unique(wDT$CellLine[wDT$ZPrimeRobustLoess>=0])`  

`r unique(wDT$Barcode[wDT$ZPrimeRobustLoess>=0])`

The cell lines and plates passing QA with SSMD values less than or equal to -3 are:
`r unique(wDT$CellLine[wDT$SSMDLoess <= -3])`  
`r unique(wDT$Barcode[wDT$SSMDLoess <= -3])`  

The cell lines and plates passing QA with SSMD values less than or equal to -2 are:
`r unique(wDT$CellLine[wDT$SSMDLoess <= -2])`  
`r unique(wDT$Barcode[wDT$SSMDLoess <= -2])`  


```{r plotQAMetrics, fig.width=4, fig.height=4}

DT <- unique(wDT[,list(Barcode,CellLine,Plate,ZPrimeRobust,SSMD,ZPrimeRobustLoess,SSMDLoess)])

p <- ggplot(DT, aes(x=factor(CellLine), y=ZPrimeRobust, shape=factor(Plate)))+
  geom_point(size = 5)+
  coord_cartesian(ylim=c(-5,1))+
  geom_hline(yintercept = 1)+geom_hline(yintercept = .5)+
  #guides(colour=guide_legend(title = "Cell Line",ncol=2), shape=guide_legend(title = "Plate"))+
  guides(shape=FALSE)+
  xlab("")+ylab("Robust Z Prime")+
  ggtitle(paste("Robust Z Prime Factors"))+
  theme(strip.text = element_text(size = rel(.5)), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),plot.title = element_text(size = rel(.8)),legend.text=element_text(size = rel(.6)),legend.title=element_text(size = rel(.6)), strip.text = element_text(size = rel(.4)))
print(p)


p <- ggplot(DT, aes(x=factor(CellLine), y=ZPrimeRobustLoess, shape=factor(Plate)))+
  geom_point(size = 5)+
  coord_cartesian(ylim=c(-5,1))+
  geom_hline(yintercept = 1)+geom_hline(yintercept = .5)+
  #guides(colour=guide_legend(title = "Cell Line",ncol=2), shape=guide_legend(title = "Plate"))+
  guides(shape=FALSE)+
  xlab("")+ylab("Robust Z Prime")+
  ggtitle(paste("Robust Z Prime Factors from Loess Normalized Well Cell Count"))+
  theme(strip.text = element_text(size = rel(.5)), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),plot.title = element_text(size = rel(.8)),legend.text=element_text(size = rel(.6)),legend.title=element_text(size = rel(.6)), strip.text = element_text(size = rel(.4)))
print(p)


p <- ggplot(DT, aes(x=factor(CellLine), y=SSMD, shape=factor(Plate)))+
  geom_point(size = 5)+
  coord_cartesian(ylim=c(-15,1))+
  geom_hline(yintercept = -3)+
  guides(shape=FALSE)+
  xlab("")+ylab("SSMD")+
  ggtitle("SSMD QA Values")+
  theme(strip.text = element_text(size = rel(.5)), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),plot.title = element_text(size = rel(.8)),legend.text=element_text(size = rel(.6)),legend.title=element_text(size = rel(.6)), strip.text = element_text(size = rel(.4)))
print(p)

p <- ggplot(DT, aes(x=factor(CellLine), y=SSMDLoess, shape=factor(Plate)))+
  geom_point(size = 5)+
  coord_cartesian(ylim=c(-15,1))+
  geom_hline(yintercept = -3)+
  guides(shape=guide_legend(title = "Plate",override.aes = list(size=2)))+
  xlab("")+ylab("SSMD")+
  ggtitle("SSMD QA Values from Loess Normalized Well Cell Count")+
  theme(strip.text = element_text(size = rel(.5)), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),plot.title = element_text(size = rel(.8)), legend.text=element_text(size = rel(.4)),legend.title=element_text(size = rel(.4)), legend.position="bottom", strip.text = element_text(size = rel(.4)))
print(p)

```



##Supplemental Material  

The following plate maps and boxplots show every step of the loess normalization (span = `r span`) on every plate in the dataset. 

```{r HeatmapsWCCLoess, echo=FALSE, fig.width=4,fig.height=2, eval=TRUE}
setkey(wDT,Barcode)
for (barcode in unique(wDT$Barcode)){
  setkey(wDT,Barcode)
  
  #Remove rows from multiple fields/images of each well
  colNames <- grep("Omero|Image|Field",colnames(wDT),value=TRUE,invert=TRUE)
  DT <- unique(wDT[barcode,colNames, with=FALSE], by=NULL)
  
  p <- ggplot(DT, aes(x=Column, y=Row, colour=WellCellCount))+
    geom_point(size=2)+
    ylim(16,1)+
    scale_x_continuous(breaks=c(1:24))+
    scale_colour_gradient(low = "white", high = "red")+
    guides(colour = guide_legend("Well Cell Count", keywidth = .5, keyheight = .5))+
    ggtitle(paste("Well Cell Count for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=factor(GeneSymbol), y=WellCellCount))+
    geom_boxplot()+
    ggtitle(paste("Well Cell Count for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=Column, y=Row, colour=mt))+
    geom_point(size=2)+
    ylim(16,1)+
    scale_x_continuous(breaks=c(1:24))+
    scale_colour_gradient(low = "white", high = "red")+
    guides(colour = guide_legend("Well Cell Count Medians", keywidth = .5, keyheight = .5))+
    ggtitle(paste("Well Cell Count Medians for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=factor(GeneSymbol), y=mt))+
    geom_boxplot()+
    ggtitle(paste("Well Cell Count Medians for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=Column, y=Row, colour=WCCRes))+
    geom_point(size=2)+
    ylim(16,1)+
    scale_x_continuous(breaks=c(1:24))+
    scale_colour_gradient(low = "white", high = "red")+
    guides(colour = guide_legend("Well Cell Count Residuals", keywidth = .5, keyheight = .5))+
    ggtitle(paste("Well Cell Count Residuals for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=factor(GeneSymbol), y=WCCRes))+
    geom_boxplot()+
    ggtitle(paste("Well Cell Count Residuals for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=Column, y=Row, colour=WCCResFitted))+
    geom_point(size=2)+
    ylim(16,1)+
    scale_x_continuous(breaks=c(1:24))+
    scale_colour_gradient(low = "white", high = "red")+
    guides(colour = guide_legend("Well Cell Count Residuals Model", keywidth = .5, keyheight = .5))+
    ggtitle(paste("Well Cell Count Residuals Loess Model for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=factor(GeneSymbol), y=WCCResFitted))+
    geom_boxplot()+
    ggtitle(paste("Well Cell Count Residuals Model for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=Column, y=Row, colour=WellCellCountLoessNorm))+
    geom_point(size=2)+
    ylim(16,1)+
    scale_x_continuous(breaks=c(1:24))+
    scale_colour_gradient(low = "white", high = "red")+
    guides(colour = guide_legend("Well Cell Count (norm)", keywidth = .5, keyheight = .5))+
    ggtitle(paste("Loess Normalized Well Cell Count for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=factor(GeneSymbol), y=WellCellCountLoessNorm))+
    geom_boxplot()+
    ggtitle(paste("Loess Normalzied Well Cell Count for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
}

```


###Explore Span Values

The following plots show the results of loess normalizing at various span values.  

```{r loessNormSpanRange, fig.width=4, fig.height=4, eval=TRUE}

for(span in c(.02, .05, .1,1)){
  colNames <- grep("Omero|Image|Field",colnames(wDT),value=TRUE,invert=TRUE)
  DT <- unique(wDT[,colNames, with=FALSE], by = NULL)
  DT <- loessNormalize(wDT, signalNames="WellCellCount", span=span)
  DT <- DT[,ZPrimeRobustLoess := calcZPrimeLoess(.SD, method="robust"), by="Barcode", .SDcols=c("GeneSymbol","WellCellCountLoessNorm")]
  DT <- DT[,SSMDLoess := SSMDPlateLoess(.SD), by="Barcode", .SDcols=c("GeneSymbol","WellCellCountLoessNorm")]
  DT <- unique(DT[,list(Barcode,CellLine,Plate,ZPrimeRobustLoess,SSMDLoess)])
  
  p <- ggplot(DT, aes(x=factor(CellLine), y=ZPrimeRobustLoess, shape=factor(Plate)))+
    geom_point(size = 5)+
    coord_cartesian(ylim=c(-5,1))+
    geom_hline(yintercept = 1)+geom_hline(yintercept = .5)+
    #guides(colour=guide_legend(title = "Cell Line",ncol=2), shape=guide_legend(title = "Plate"))+
    guides(shape=FALSE)+
    xlab("")+ylab("Robust Z Prime")+
    ggtitle(paste("Robust Z Prime Factors from Loess Normalized Well Cell Count"))+
    theme(strip.text = element_text(size = rel(.5)), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),plot.title = element_text(size = rel(.8)),legend.text=element_text(size = rel(.6)),legend.title=element_text(size = rel(.6)), strip.text = element_text(size = rel(.4)))
  print(p)
  
  
  p <- ggplot(DT, aes(x=factor(CellLine), y=SSMDLoess, shape=factor(Plate)))+
    geom_point(size = 5)+
    coord_cartesian(ylim=c(-15,1))+
    geom_hline(yintercept = -3)+
    guides(shape=guide_legend(title = "Plate",override.aes = list(size=2)))+
    xlab("")+ylab("SSMD")+
    ggtitle("SSMD QA Values from Loess Normalized Well Cell Count")+
    theme(strip.text = element_text(size = rel(.5)), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),plot.title = element_text(size = rel(.8)), legend.text=element_text(size = rel(.4)),legend.title=element_text(size = rel(.4)), legend.position="bottom", strip.text = element_text(size = rel(.4)))
  print(p)
  
  
  cat(paste("Plates with Z Prime Robust factor >= 0 on Loess Normed WCC =",sum(DT$ZPrimeRobustLoess>=0, na.rm=TRUE),"for span=",span,"\n"))
  cat(paste("Plates with SSMD <= -3 on Loess Normed WCC =",sum(DT$SSMDLoess<= -3, na.rm=TRUE),"for span=",span,"\n"))
  cat(paste("Plates with SSMD <= -2 on Loess Normed WCC =",sum(DT$SSMDLoess<= -2, na.rm=TRUE),"for span=",span,"\n"))
  
  
}
```
