---
title: "Atwaters siRNA Normalization Analysis"
output: html_document
date: '`r Sys.Date()`'
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,
                      echo=FALSE, warning=FALSE, message=FALSE)
```

```{r setup}
library("ggplot2")
library("data.table")
library("MEMA")
library("grid")
library("knitr")
library("gplots")
library("RColorBrewer")
library("DT")
library("d3heatmap")
library("plotly")
library(scales)

source("WellPlateFunctions.R")

#Calculate SSMD
SSMDPlate <- function(df){
  if(length(unique(df)))
    mup <- mean(df$WellCellCount[df$GeneSymbol=="PLK1"], na.rm = TRUE)
  sdp <- sd(df$WellCellCount[df$GeneSymbol=="PLK1"], na.rm = TRUE)
  mun <- mean(df$WellCellCount[df$GeneSymbol=="NegCtrl"], na.rm = TRUE)
  sdn <- sd(df$WellCellCount[df$GeneSymbol=="NegCtrl"], na.rm = TRUE)
  SSMD <- (mup-mun)/sqrt(sdp^2+sdn^2)
}



calcZPrimeLoess <- function(dt, method="original", posCtrl="PLK1"){
  imageHTS::zprime(dt[GeneSymbol=="NegCtrl",WellCellCountLoessNorm],dt[GeneSymbol==posCtrl,WellCellCountLoessNorm], method)
}

wDT <- fread("AtwatersWell_v1.1.txt")
wDT$OmeroDetailURL <- paste0('<a href="https://omero.ohsu.edu/webclient/img_detail/',wDT$ImageID,'/"',' target="_blank">Omero</a>')
wDT$OmeroThumbnailURL <- paste0('<a href="https://omero.ohsu.edu/webclient/render_thumbnail/',wDT$ImageID,'/"',' target="_blank">Omero</a>')
wDT$OmeroImageURL <- paste0('<a href="https://omero.ohsu.edu/webclient/render_image/',wDT$ImageID,'/"',' target="_blank">Omero</a>')

repDT <- fread("AtwatersRep_v1.1.txt",showProgress=FALSE)
barcodes <- unique(repDT$Barcode)[1]
```
```{r filter, echo=TRUE}
#Manually filter out selected rows of poor quality
# AT384P20038 (ZR75B): rows C and D have very low cell counts
# AT384P10037 (ZR75B): rows C and D have very low cell counts
# AT384P10049 (HCC38): columns 12 and 13 have missing cells
wDT <- wDT[!(grepl("AT384P10037",wDT$Barcode) & grepl("^3$|^4$",wDT$Row)),]
wDT <- wDT[!(grepl("AT384P20038",wDT$Barcode) & grepl("^3$|^4$",wDT$Row)),]
wDT <- wDT[!(grepl("AT384P10049",wDT$Barcode) & grepl("^12$|^13$",wDT$Column)),]

```


```{r loessNorm}

span=.05

wDT <- loessNormalize(wDT, signalNames="WellCellCount", span=span)

```

###Loess Normalization

The Well Cell Count is loess normalized within each plate using the following method:

Calculate the m~target~ values which are the median Well Cell Counts of the replicates of each siRNA or control  
Calculate the residuals by subtracting m~target~ from each Well Cell Count value  
Fit a loess model to the residuals  
At each well, subtract the fitted loess value from the Well Cell Count  

The loess span parameter is set to `r span` for this run of the analysis.

```{r PLK1controlsAnalysis,  fig.width=8, fig.height=6, eval=FALSE}


colNames <- grep("Omero|Image|Field",colnames(wDT),value=TRUE,invert=TRUE)
wnfDT <- unique(wDT[,colNames, with=FALSE])
setkey(wnfDT,GeneSymbol)
wnfDT <- wnfDT[c("PLK1","NegCtrl")]
wnfDT <- wnfDT[,ZPrimeOriginal := calcZPrime(.SD, method="original"), by="Barcode", .SDcols=c("GeneSymbol","WellCellCount")]
wnfDT <- wnfDT[,ZPrimeRobust := calcZPrime(.SD, method="robust"), by="Barcode", .SDcols=c("GeneSymbol","WellCellCount")]
wnfDT <- wnfDT[,ZPrimeFixsd := calcZPrime(.SD, method="fixsd"), by="Barcode", .SDcols=c("GeneSymbol","WellCellCount")]
wnfDT <- wnfDT[,SSMD := SSMDPlate(.SD), by="Barcode", .SDcols=c("GeneSymbol","WellCellCount")]
wnfDT <- wnfDT[,SSMDLoess := SSMDPlate(.SD), by="Barcode", .SDcols=c("GeneSymbol","WellCellCountLoessNorm")]
colNames <- c("Barcode", "CellLine","Plate", "ZPrimeRobust","ZPrimeOriginal","ZPrimeFixsd","ZPrimeFixsd","ZPrimeRobustLoess","SSMD","SSMDLoess")

wnfDT <- wnfDT[,ZPrimeRobustLoess := calcZPrimeLoess(.SD, method="robust"), by="Barcode", .SDcols=c("GeneSymbol","WellCellCountLoessNorm")]

DT <- unique(wnfDT[,colNames, with=FALSE])


p <- ggplot(DT, aes(x=factor(CellLine), y=ZPrimeRobust, colour=CellLine, shape=factor(Plate)))+
  geom_point(size = 5)+
  coord_cartesian(ylim=c(-5,1))+
  geom_hline(yintercept = 1)+geom_hline(yintercept = .5)+
  guides(colour=guide_legend(title = "Cell Line",ncol=2), shape=guide_legend(title = "Plate"))+
  xlab("")+ylab("Robust Z Prime")+
  ggtitle("Robust Z Prime Factors")+
  theme(strip.text = element_text(size = rel(.5)), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),plot.title = element_text(size = rel(1.5)),legend.text=element_text(size = rel(.6)),legend.title=element_text(size = rel(.6)), strip.text = element_text(size = rel(.4)))
print(p)

p <- ggplot(DT, aes(x=factor(CellLine), y=ZPrimeRobustLoess, colour=CellLine, shape=factor(Plate)))+
  geom_point(size = 5)+
  coord_cartesian(ylim=c(-5,1))+
  geom_hline(yintercept = 1)+geom_hline(yintercept = .5)+
  guides(colour=guide_legend(title = "Cell Line",ncol=2), shape=guide_legend(title = "Plate"))+
  xlab("")+ylab("Robust Z Prime")+
  ggtitle("Robust Z Prime Factors on Loess Normed WCC")+
  theme(strip.text = element_text(size = rel(.5)), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),plot.title = element_text(size = rel(1.5)),legend.text=element_text(size = rel(.6)),legend.title=element_text(size = rel(.6)), strip.text = element_text(size = rel(.4)))
print(p)

p <- ggplot(DT, aes(x=factor(CellLine), y=SSMD, colour=CellLine, shape=factor(Plate)))+
  geom_point(size = 5)+
  coord_cartesian(ylim=c(-15,1))+
  geom_hline(yintercept = -3)+
  guides(colour=guide_legend(title = "Cell Line",ncol=2), shape=guide_legend(title = "Plate"))+
  xlab("")+ylab("SSMD")+
  ggtitle("SSMD QA Values")+
  theme(strip.text = element_text(size = rel(.5)), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),plot.title = element_text(size = rel(1.5)),legend.text=element_text(size = rel(.6)),legend.title=element_text(size = rel(.6)), strip.text = element_text(size = rel(.4)))
print(p)


p <- ggplot(DT, aes(x=factor(CellLine), y=SSMDLoess, colour=CellLine, shape=factor(Plate)))+
  geom_point(size = 5)+
  coord_cartesian(ylim=c(-15,1))+
  geom_hline(yintercept = -3)+
  guides(colour=guide_legend(title = "Cell Line",ncol=2), shape=guide_legend(title = "Plate"))+
  xlab("")+ylab("SSMD")+
  ggtitle("SSMD QA Values on Loess Normalized WCC")+
  theme(strip.text = element_text(size = rel(.5)), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),plot.title = element_text(size = rel(1.5)),legend.text=element_text(size = rel(.6)),legend.title=element_text(size = rel(.6)), strip.text = element_text(size = rel(.4)))
print(p)



```

```{r HeatmapsWCCLoess, echo=FALSE, fig.width=4,fig.height=2, eval=TRUE}
setkey(wDT,Barcode)
for (barcode in unique(wDT$Barcode)){
  setkey(wDT,Barcode)
  DT <- wDT[barcode]
  
  p <- ggplot(DT, aes(x=Column, y=Row, colour=WellCellCount))+
    geom_point(size=2)+
    ylim(16,1)+
    scale_x_continuous(breaks=c(1:24))+
    scale_colour_gradient(low = "white", high = "red")+
    guides(colour = guide_legend("Well Cell Count", keywidth = .5, keyheight = .5))+
    ggtitle(paste("Well Cell Count for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=factor(GeneSymbol), y=WellCellCount))+
    geom_boxplot()+
    ggtitle(paste("Well Cell Count for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=Column, y=Row, colour=mt))+
    geom_point(size=2)+
    ylim(16,1)+
    scale_x_continuous(breaks=c(1:24))+
    scale_colour_gradient(low = "white", high = "red")+
    guides(colour = guide_legend("Well Cell Count Medians", keywidth = .5, keyheight = .5))+
    ggtitle(paste("Well Cell Count Medians for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=factor(GeneSymbol), y=mt))+
    geom_boxplot()+
    ggtitle(paste("Well Cell Count Medians for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=Column, y=Row, colour=WCCRes))+
    geom_point(size=2)+
    ylim(16,1)+
    scale_x_continuous(breaks=c(1:24))+
    scale_colour_gradient(low = "white", high = "red")+
    guides(colour = guide_legend("Well Cell Count Residuals", keywidth = .5, keyheight = .5))+
    ggtitle(paste("Well Cell Count Residuals for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=factor(GeneSymbol), y=WCCRes))+
    geom_boxplot()+
    ggtitle(paste("Well Cell Count Residuals for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=Column, y=Row, colour=WCCResFitted))+
    geom_point(size=2)+
    ylim(16,1)+
    scale_x_continuous(breaks=c(1:24))+
    scale_colour_gradient(low = "white", high = "red")+
    guides(colour = guide_legend("Well Cell Count Residuals Model", keywidth = .5, keyheight = .5))+
    ggtitle(paste("Well Cell Count Residuals Loess Model for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=factor(GeneSymbol), y=WCCResFitted))+
    geom_boxplot()+
    ggtitle(paste("Well Cell Count Residuals Model for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=Column, y=Row, colour=WellCellCountLoessNorm))+
    geom_point(size=2)+
    ylim(16,1)+
    scale_x_continuous(breaks=c(1:24))+
    scale_colour_gradient(low = "white", high = "red")+
    guides(colour = guide_legend("Well Cell Count (norm)", keywidth = .5, keyheight = .5))+
    ggtitle(paste("Loess Normalized Well Cell Count for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
  p <- ggplot(DT, aes(x=factor(GeneSymbol), y=WellCellCountLoessNorm))+
    geom_boxplot()+
    ggtitle(paste("Loess Normalzied Well Cell Count for",unique(DT$CellLine), "cells in plate",unique(DT$Barcode)))+
    xlab("")+ylab("")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=rel(.5)),plot.title = element_text(size = rel(.6)),legend.text=element_text(size = rel(.3)),legend.title=element_text(size = rel(.3)))
  suppressWarnings(print(p))
  
}

```


```{r rawZprimeSSMD}

  colNames <- grep("Omero|Image|Field",colnames(wDT),value=TRUE,invert=TRUE)
  DT <- unique(wDT[,colNames, with=FALSE])
  DT <- DT[,ZPrimeRobust := calcZPrime(.SD, method="robust"), by="Barcode", .SDcols=c("GeneSymbol","WellCellCount")]
  DT <- DT[,SSMD := SSMDPlate(.SD), by="Barcode", .SDcols=c("GeneSymbol","WellCellCount")]
  DT <- unique(DT[,list(Barcode,CellLine,Plate,ZPrimeRobust,SSMD)])


p <- ggplot(DT, aes(x=factor(CellLine), y=ZPrimeRobust, colour=CellLine, shape=factor(Plate)))+
  geom_point(size = 5)+
  coord_cartesian(ylim=c(-5,1))+
  geom_hline(yintercept = 1)+geom_hline(yintercept = .5)+
  guides(colour=guide_legend(title = "Cell Line",ncol=2), shape=guide_legend(title = "Plate"))+
  xlab("")+ylab("Robust Z Prime")+
  ggtitle("Robust Z Prime Factors")+
  theme(strip.text = element_text(size = rel(.5)), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),plot.title = element_text(size = rel(1.5)),legend.text=element_text(size = rel(.6)),legend.title=element_text(size = rel(.6)), strip.text = element_text(size = rel(.4)))
print(p)


p <- ggplot(DT, aes(x=factor(CellLine), y=SSMD, colour=CellLine, shape=factor(Plate)))+
  geom_point(size = 5)+
  coord_cartesian(ylim=c(-15,1))+
  geom_hline(yintercept = -3)+
  guides(colour=guide_legend(title = "Cell Line",ncol=2), shape=guide_legend(title = "Plate"))+
  xlab("")+ylab("SSMD")+
  ggtitle("SSMD QA Values")+
  theme(strip.text = element_text(size = rel(.5)), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),plot.title = element_text(size = rel(1.5)),legend.text=element_text(size = rel(.6)),legend.title=element_text(size = rel(.6)), strip.text = element_text(size = rel(.4)))
print(p)

cat(paste("Plates with Z Prime Robust factor >= 0 =",sum(DT$ZPrimeRobust>=0),"\n"))
cat(paste("Plates with SSMD <= -3 =",sum(DT$SSMD <= -3, na.rm=TRUE),"\n"))

```

###Explore Span Values



```{r loessNormSpanRange}

for(span in c(.02, .05, .1,1)){
  colNames <- grep("Omero|Image|Field",colnames(wDT),value=TRUE,invert=TRUE)
  DT <- unique(wDT[,colNames, with=FALSE])
  DT <- loessNormalize(wDT, signalNames="WellCellCount", span=span)
  DT <- DT[,ZPrimeRobust := calcZPrime(.SD, method="robust"), by="Barcode", .SDcols=c("GeneSymbol","WellCellCount")]
  DT <- DT[,ZPrimeRobustLoess := calcZPrimeLoess(.SD, method="robust"), by="Barcode", .SDcols=c("GeneSymbol","WellCellCountLoessNorm")]
  DT <- DT[,SSMD := SSMDPlate(.SD), by="Barcode", .SDcols=c("GeneSymbol","WellCellCount")]
  DT <- DT[,SSMDLoess := SSMDPlate(.SD), by="Barcode", .SDcols=c("GeneSymbol","WellCellCountLoessNorm")]
  DT <- unique(DT[,list(Barcode,CellLine,Plate,ZPrimeRobust,ZPrimeRobustLoess,SSMD,SSMDLoess)])
  
 
  p <- ggplot(DT, aes(x=factor(CellLine), y=ZPrimeRobustLoess, colour=CellLine, shape=factor(Plate)))+
    geom_point(size = 5)+
    coord_cartesian(ylim=c(-5,1))+
    geom_hline(yintercept = 1)+geom_hline(yintercept = .5)+
    guides(colour=guide_legend(title = "Cell Line",ncol=2), shape=guide_legend(title = "Plate"))+
    xlab("")+ylab("Robust Z Prime")+
    ggtitle(paste("Robust Z Prime Factors on Loess Normed WCC, Span=",span))+
    theme(strip.text = element_text(size = rel(.5)), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),plot.title = element_text(size = rel(1.5)),legend.text=element_text(size = rel(.6)),legend.title=element_text(size = rel(.6)), strip.text = element_text(size = rel(.4)))
  print(p)
  
  
  p <- ggplot(DT, aes(x=factor(CellLine), y=SSMDLoess, colour=CellLine, shape=factor(Plate)))+
    geom_point(size = 5)+
    coord_cartesian(ylim=c(-15,1))+
    geom_hline(yintercept = -3)+
    guides(colour=guide_legend(title = "Cell Line",ncol=2), shape=guide_legend(title = "Plate"))+
    xlab("")+ylab("SSMD")+
    ggtitle(paste("SSMD QA Values on Loess Normalized WCC, Span=",span))+
    theme(strip.text = element_text(size = rel(.5)), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),plot.title = element_text(size = rel(1.5)),legend.text=element_text(size = rel(.6)),legend.title=element_text(size = rel(.6)), strip.text = element_text(size = rel(.4)))
  print(p)
  cat(paste("Plates with Z Prime Robust factor >= 0 on Loess Normed WCC =",sum(DT$ZPrimeRobustLoess>=0),"for span=",span,"\n"))
  cat(paste("Plates with SSMD <= -2 on Loess Normed WCC =",sum(DT$SSMDLoess<= -2),"for span=",span,"\n"))
cat(paste("Plates with SSMD <= -3 on Loess Normed WCC =",sum(DT$SSMDLoess<= -3),"for span=",span,"\n"))

}
```
